Abstract

### 2025-10-09 output/Response_to_action_editor_reviewers-2.Rmd

###### Then one could make another version of the left panel with contours for the rho=0.5 population distribution. Comparison of that with the previously-described version would show that the arrows representing discrimination do not have the same numerical meaning with respect to the variance (standard deviation) of the population distribution *in the direction of the arrows*, which is part of the point of all of this.

###### Two graphics with orthogonal axes but different population distributions are the way orthogonal vs. oblique solutions are represented in MIRT analyses. The ms. misses an opportunity to clarify that. Different population distributions on orthogonal axes are the way MIRT folks have thought about factor correlation; factor analysts have thought about 'rotation' and oblique axes (the left panel of Figure 1); it's two sides of the same coin.

This is indeed the message we want to convey with our manuscript, and are thankful to the reviewer for the comments.
We have taken into account the suggestion and added the contours plots accordingly.
This is represented as panel (b) in the new version.
The text also explains the different interpretation of the coordinates relative to the distribution now.
Please see the new paragraph added in third place in section "Graphical Representation Example" (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):

> Panels (a) and (b) show that,
> even if the agnostic coordinates are the same in both cases,
> their relative position with respect to the latent trait distribution is not
> ---this is the effect of disregarding the covariance structure.
> The correlation-based coordinates, plotted in panels (c) and (d),
> are also different in general from their agnostic counterpart (panel (b)).
> Although they do not vary in some cases (items 1 and 4),
> the difference is especially salient in the location of items 2 and 3,
> which are close to the 90% contour in the agnostic version
> but almost on the 50% contour in the correlation-based one. [...]

@@

- The second paragraph in section "Graphical Representation" now clarifies better that the representation in the test space must use its basis instead of the canonical basis, to make the plot be a faithful representation of the actual directions of the items with respect to the latent trait dimensions.
This paragraph now reads (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):

> [...] Coordinates in a rectangular Cartesian system are commonly assumed
> to be simply represented in the canonical basis.
> To represent an item in the test space properly, however,
> we must use its basis $`r TEST_SPACE_BASIS`$,
> which implies pre-multiplying its coordinates
> by $`r TRANSFORM_MATRIX_TRANSP_INV`$
> to find the equivalent coordinates in the canonical basis. [...]

- The section "Graphical Representation Example" has been expanded to explain better both Table 1 and Figure 1.
Starting in the first paragraph (p. <!-- # TODO: Complete -->, l. <!-- # TODO: Complete -->), this section now reads:

> [...] The example in Figure 1 showcases a representation of
> a set of items with two different latent trait distributions.
> (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
> for simplicity, but note that they represent the coordinates of
> a latent trait point expressed in test space coordinates,
> that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
> Their parameters are shown in Table 1,
> both in the original `r MODEL_ACRONYM` metric and
> in the multidimensional metric under the two cases,
> which differ in the latent trait correlation.
> In the first case (columns $`r CORR_ORTH_OUT`$),
> the two latent dimensions are independent;
> in the second one ($`r CORR_OBL_OUT`$),
> the correlation between them is .5.
> The same `r MODEL_ACRONYM` parameters are used in both cases,
> but their (correlation-based) multidimensional parameters change
> with the correlation.
> If we used the agnostic version instead, the multidimensional parameters,
> given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.
> 
> Panel (a) of Figure 1 shows
> the item representation in the uncorrelated case,
> while the $`r CORR_OBL_OUT`$ case is shown in panels (b) to (d)
> under different conditions:
> with item coordinates computed from the agnostic version of the parameters
> (panel (b)), or from the correlation-based version,
> either using an appropriate oblique basis (panel (c)),
> or plotting them directly in the canonical basis (panel (d)).
> Keeping the horizontal axis invariant in panel (c)
> allows for a better comparison with the other plots;
> therefore, we have chosen a $`r TRANSFORM_MATRIX_TRANSP_INV`$ that
> only rotates the vertical axis [@harman_modern_1976]:
> 
> \begin{equation}
>   `r TRANSFORM_MATRIX_EXAMPLE_EQ`.
> \end{equation}
> 
> In all four panels, the contours of a bivariate standard normal density
> with the corresponding correlation are plotted along with the items
> (from outer to inner, the contours are the `r contour_breaks_out` of
> the density at its mode).
> The stretched ellipses in panel (c) are the result of transforming
> the distribution in the latent space to the test space,
> according to Equation \@ref(eq:trait-coords-spaces-transform).
> This implies that the covariance matrix of the distribution projected into
> the test space is transformed inversely as in the latent space, i.e.,
> $`r COV_MATRIX_TRANSF_TS_EQ`$.
> 
> Panels (a) and (b) show that,
> even if the agnostic coordinates are the same in both cases,
> their relative position with respect to the latent trait distribution is not
> ---this is the effect of disregarding the covariance structure.
> The correlation-based coordinates, plotted in panels (c) and (d),
> are also different in general from their agnostic counterpart (panel (b)).
> Although they do not vary in some cases (items 1 and 4),
> the difference is especially salient in the location of items 2 and 3,
> which are close to the 90% contour in the agnostic version
> but almost on the 50% contour in the correlation-based one.
> Panel (d) also illustrates the effect of plotting the coordinates directly
> on the canonical basis;
> the item locations and their positions relative to the distribution
> are correctly represented (which makes comparing with panel (b) easier),
> but their discrimination parameters and angles with the axes are distorted:
> For example, items 4 and 5 seem to be equally discriminating,
> when Table 1 panel (c) show they are not;
> and item 1 forms 45-degree angles with both axes,
> instead of the 30-degree ones shown in Table 1.
> 
> A comparison of panels (a) and (c) allows understanding the effect of
> a positive correlation on the $`r MDISC_CORR_PARAM`$
> (and consequently the $`r MIL_CORR_PARAM`$) value: [...]

Finally, we have added a fourth panel (panel (d)) that illustrates how the items would be plotted if we omitted using the proper test space basis (and just using the canonical basis instead).
Incidentally, this also serves the purpose of allowing an easier comparison of the coordinates computed from the two versions (the agnostic, and the correlation-based one) of the multidimensional parameters.
We have briefly explained this in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete --> (please refer to the changes explained above).

### 2025-10-09 output/Main_text.Rmd
Correct the grammar of "plot" in simple past
Should be "plotted" instead

In all four panels, the contours of a bivariate standard normal density
with the corresponding correlation are plot along with the items
(from outer to inner, the contours are the `r contour_breaks_out` of
the density at its mode).
The stretched ellipses in panel (c) are the result of transforming
the distribution in the latent space to the test space,
according to Equation \@ref(eq:trait-coords-spaces-transform).
This implies that the covariance matrix of the distribution projected into
the test space is transformed inversely as in the latent space, i.e.,
$`r COV_MATRIX_TRANSF_TS_EQ`$.

Panels (a) and (b) show that,
even if the agnostic coordinates are the same in both cases,
their relative position with respect to the latent trait distribution is not
---this is the effect of disregarding the covariance structure.
The correlation-based coordinates, plot in panels (c) and (d),
are also different in general from their agnostic counterpart (panel (b)).
Although they do not vary in some cases (items 1 and 4),
the difference is especially salient in the location of items 2 and 3,
which are close to the 90% contour in the agnostic version
but almost on the 50% contour in the correlation-based one.
Panel (d) also illustrates the effect of plotting the coordinates directly
on the canonical basis;
the item locations and their positions relative to the distribution
are correctly represented (which makes comparing with panel (b) easier),
but their discrimination parameters and angles with the axes are distorted:
For example, items 4 and 5 seem to be equally discriminating,
when Table \@ref(tab:example-items-table-out) panel (c) show they are not;
and item 1 forms 45-degree angles with both axes,
instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).

@@

```{r item-plot-out}
#| cache:          false
#| fig.cap:        |
#|   Item vector plots with uncorrelated (a) and correlated latent dimensions.
#|   The latter coordinates are computed with
#|   (b) the agnostic multidimensional parameters, or
#|   with the covariance/correlation-based ones, which are then plot either
#|   (c) in an oblique basis, appropriate for the covariance structure, or
#|   (d) in the canonical basis.
#| fig.topcaption: true
#| fig.height:     4.7
#| fig.width:      6.65
#| warning:        false
(plot_orth_uncorr_contours + plot_orth_corr) /
  (plot_oblique_ts + plot_cov_based_orth) +
  plot_layout(guides = "collect")

>>>>

In all four panels, the contours of a bivariate standard normal density
with the corresponding correlation are plotted along with the items
(from outer to inner, the contours are the `r contour_breaks_out` of
the density at its mode).
The stretched ellipses in panel (c) are the result of transforming
the distribution in the latent space to the test space,
according to Equation \@ref(eq:trait-coords-spaces-transform).
This implies that the covariance matrix of the distribution projected into
the test space is transformed inversely as in the latent space, i.e.,
$`r COV_MATRIX_TRANSF_TS_EQ`$.

Panels (a) and (b) show that,
even if the agnostic coordinates are the same in both cases,
their relative position with respect to the latent trait distribution is not
---this is the effect of disregarding the covariance structure.
The correlation-based coordinates, plotted in panels (c) and (d),
are also different in general from their agnostic counterpart (panel (b)).
Although they do not vary in some cases (items 1 and 4),
the difference is especially salient in the location of items 2 and 3,
which are close to the 90% contour in the agnostic version
but almost on the 50% contour in the correlation-based one.
Panel (d) also illustrates the effect of plotting the coordinates directly
on the canonical basis;
the item locations and their positions relative to the distribution
are correctly represented (which makes comparing with panel (b) easier),
but their discrimination parameters and angles with the axes are distorted:
For example, items 4 and 5 seem to be equally discriminating,
when Table \@ref(tab:example-items-table-out) panel (c) show they are not;
and item 1 forms 45-degree angles with both axes,
instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).

@@

```{r item-plot-out}
#| cache:          false
#| fig.cap:        |
#|   Item vector plots with uncorrelated (a) and correlated latent dimensions.
#|   The latter coordinates are computed with
#|   (b) the agnostic multidimensional parameters, or
#|   with the correlation-based ones, which are then plotted either
#|   (c) in an oblique basis, appropriate for the covariance structure, or
#|   (d) in the canonical basis.
#| fig.topcaption: true
#| fig.height:     4.7
#| fig.width:      6.65
#| warning:        false
(plot_orth_uncorr_contours + plot_orth_corr) /
  (plot_oblique_ts + plot_cov_based_orth) +
  plot_layout(guides = "collect")
